Storm surges in the Strait of Georgia simulated with a regional model

### 2015-01-02 AO_guides/manuscriptAO.tex
modified title, author order and added webtide website

\title{Storm surge modelling in the Strait of Georgia}

\author{Nancy Soontiens$^{1}$$^{\ast}$\thanks{$^\ast$Corresponding author's email: nsoontie@eos.ubc.ca
\vspace{6pt}}, Susan Allen$^{1}$,  Doug Latornell$^{1}$,  Kate Le Sou\"{e}f$^{1}$, Idalia Machuca$^{1}$ \\ 
Youyu Lu$^{2}$, Keith Thompson$^{3}$, Jean-Philippe Paquin$^{3}$, Vasily Korabel$^{3}$
\\\vspace{6pt}  $^{1}${\em{University of British Columbia}};
$^{2}${\em{Bedford Insitute of Oceanography}}; 
$^{3}${\em{Dalhousie University}}\\\received{v1.2 released April 2014} }

@@

\section{Model Configuration}\label{sec:config}

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on validating the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation.

@@

The model includes two open boundaries that connect to the Pacific Ocean, the western boundary of the Strait of Juan de Fuca as well as Johnstone Strait at the north, both of which are forced with eight tidal constituents (K$_1$ ,O$_1$, P$_1$, Q$_1$, M$_2$, K$_2$, N$_2$, S$_2$), temperature and salinity climatologies, and the sea surface height anomaly. Tidal heights and currents at grid points along the Juan de Fuca boundary were extracted from Webtide, an online web prediction model for the northeast Pacific Ocean, which is based on work by \citet{foreman2000webtide}. The Johnstone Strait boundary was forced with current and elevation tidal harmonics measured and calculated by \citet{thomson1980johnstone} for the major M$_2$ and K$_1$ constituents. Additionally, O$_1$ and S$_2$ elevation harmonics from their measurements were employed. The remaining constituents were extrapolated from Webtide. At the western boundary, the tidal elevation and currents forcing parameters were tuned for a good match with observations at locations within the Strait of Georgia, including Point Atkinson, Gibsons Landing, Winchelsea and Halfmoon Bay. At the northern boundary, the tides were tuned for good matches at Yorke Island and Kelsey Bay. Further details on the tuning are provided in Appendix \ref{sec:appendix}.

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results from a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. %other references?

Sea surface height at the mouth of Juan de Fuca was set using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain.  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. Tidal predictions were generated with a tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year prior to the surge event to avoid over predicted constituents during high surge years. The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait. The tidal forcing and sea surface height was applied to the barotropic velocity forcing with the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4, the version used here.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. At the north baroclinic velocities were set to zero.

@@

\section{Discussion}\label{sec:diss}
In this article, we have configured a three-dimensional numerical ocean model of the Salish Sea for accurate hindcasts of storm surges in the Strait of Georgia. The model is forced with eight tidal constituents at the western and northern open boundaries, as well as sea surface height anomaly from de-tided observations at Tofino and Port Hardy. Additionally, wind stress and an inverse barometer effect were forced at the ocean surface using an atmospheric reforecasting product called CGRF \citep{smith2014new}. Storm surge simulations began with zero initial conditions for the currents and sea surface height and were commenced several days prior to the event of interest to spin up the velocity fields.

@@

\appendices
\section{Appendix}\label{sec:appendix}
This section details the procedure used to tune the tidal forcing for a good match with observations in the Strait of Georgia. Originally, the eight most important tidal constituents were extracted from a tide prediction tool called Webtide \citep{foreman2000webtide} at the western Juan de Fuca boundary. At the northern boundary,  M$_2$ and K$_1$ tidal harmonics for both elevation and currents were taken from \citet{thomson1980johnstone}, as well as elevation harmonics for O$_1$ and S$_2$. The remaining constituents were extrapolated from Webtide based on the ratio between S$_2$/M$_2$ amplitudes and phase differences for the semi-diurnal constituents and O$_1$/K$_1$ for the diurnal constituents. However, these forcing conditions did not result in a good match between observed and modelled tidal elevation harmonics in the Strait of Georgia, the main region of interest (although the matches in the Strait of Juan de Fuca were adequate). As such, tidal forcing was tuned, by first adjusting the western boundary M$_2$ and K$_1$ phase and K$_1$ amplitude for a good match in the Strait of Georgia (specifically at Point Atkinson, Gibsons Landing, Winchelsea and Halfmoon Bay). This tuning was performed on simulations that used only M$_2$ and K$_1$ forcing. Next, the M$_2$ and K$_1$ fluxes at the northern boundary were adjusted because the cross-sectional area of transect used by \citet{thomson1980johnstone} to measure tidal currents in Johnstone Strait is smaller than the cross-sectional area of our northern boundary. Finally, the remaining constituents were added, starting with O$_1$, then S$_2$, and then in pairs P$_1$/N$_2$ and Q$_1$/K$_2$. Final adjustments were made by slightly varying amplitude and phase, evaluating the changes to M$_2$, K$_1$ and the ratios S$_2$/M$_2$, N$_2$/M$_2$, K$_2$/M$_2$, O$_1$/K$_1$, P$_1$/K$_1$, Q$_1$/K$_1$ and then choosing new values to move closer to observations by linear interpolation.

>>>>

\title{Storm surges in the Strait of Georgia simulated with a regional model}

\author{Nancy Soontiens$^{1}$$^{\ast}$\thanks{$^\ast$Corresponding author's email: nsoontie@eos.ubc.ca
\vspace{6pt}}, Susan Allen$^{1}$,  Doug Latornell$^{1}$,  Kate Le Sou\"{e}f$^{1}$, Idalia Machuca$^{1}$,\\ Jean-Philippe Paquin$^{2}$,
Youyu Lu$^{3}$, Keith Thompson$^{2}$, Vasily Korabel$^{2}$
\\\vspace{6pt}  $^{1}${\em{University of British Columbia}};
$^{2}${\em{Dalhousie University}};
$^{3}${\em{Bedford Insitute of Oceanography}}\\\received{v1.2 released April 2014} }

@@

\section{Model Description}\label{sec:config}

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on validating the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study.

@@

The model includes two open boundaries that connect to the Pacific Ocean, the western boundary of the Strait of Juan de Fuca as well as Johnstone Strait at the north, both of which are forced with eight tidal constituents (K$_1$ ,O$_1$, P$_1$, Q$_1$, M$_2$, K$_2$, N$_2$, S$_2$), temperature and salinity climatologies, and the sea surface height anomaly. Tidal heights and currents at grid points along the Juan de Fuca boundary were extracted from Webtide \citep{webtide}, an online web prediction model that provides tidal predictions for the northeast Pacific Ocean based on work by \citet{foreman2000webtide}. The Johnstone Strait boundary was forced with current and elevation tidal harmonics measured and calculated by \citet{thomson1980johnstone} for the major M$_2$ and K$_1$ constituents. Additionally, O$_1$ and S$_2$ elevation harmonics from their measurements were employed. The remaining constituents were extrapolated from Webtide. At the western boundary, the tidal elevation and currents forcing parameters were tuned for a good match with observations at locations within the Strait of Georgia, including Point Atkinson, Gibsons Landing, Winchelsea and Halfmoon Bay. At the northern boundary, the tides were tuned for good matches at Yorke Island and Kelsey Bay. Further details on the tuning are provided in Appendix \ref{sec:appendix}.

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results from a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid to give a two-dimensional forcing that varies with time.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. %other references?

Sea surface height was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain.  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. Tidal predictions were generated with a tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year prior to the surge event to avoid over predicted constituents during high surge years. The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait. The tidal forcing and sea surface height was applied to the barotropic velocity forcing with the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}. The barotropic velocity induced by the surge amplitude is set to zero in the Flather scheme since this data is unavailable, which is a resonable choice since the surge amplitudes are generally much smaller than the tidal range. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4, the version used here.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. At the north baroclinic velocities were set to zero.

@@

\section{Conclusion and discussion}\label{sec:diss}
In this article, we have configured a three-dimensional numerical ocean model of the Salish Sea for accurate hindcasts of storm surges in the Strait of Georgia. The model is forced with eight tidal constituents at the western and northern open boundaries, as well as sea surface height anomaly from de-tided observations at Tofino and Port Hardy. Additionally, wind stress and an inverse barometer effect were forced at the ocean surface using an atmospheric reforecasting product called CGRF \citep{smith2014new}. Storm surge simulations began with zero initial conditions for the currents and sea surface height and were commenced several days prior to the event of interest to spin up the velocity fields.

@@

\appendices
\section{Appendix}\label{sec:appendix}
This section details the procedure used to tune the tidal forcing for a good match with observations in the Strait of Georgia. Originally, the eight most important tidal constituents were extracted from a tide prediction tool called Webtide \citep{webtide} at the western Juan de Fuca boundary based on work by \citet{foreman2000webtide}. At the northern boundary,  M$_2$ and K$_1$ tidal harmonics for both elevation and currents were taken from \citet{thomson1980johnstone}, as well as elevation harmonics for O$_1$ and S$_2$. The remaining constituents were extrapolated from Webtide based on the ratio between S$_2$/M$_2$ amplitudes and phase differences for the semi-diurnal constituents and O$_1$/K$_1$ for the diurnal constituents. However, these forcing conditions did not result in a good match between observed and modelled tidal elevation harmonics in the Strait of Georgia, the main region of interest (although the matches in the Strait of Juan de Fuca were adequate). As such, tidal forcing was tuned, by first adjusting the western boundary M$_2$ and K$_1$ phase and K$_1$ amplitude for a good match in the Strait of Georgia (specifically at Point Atkinson, Gibsons Landing, Winchelsea and Halfmoon Bay). This tuning was performed on simulations that used only M$_2$ and K$_1$ forcing. Next, the M$_2$ and K$_1$ fluxes at the northern boundary were adjusted because the cross-sectional area of transect used by \citet{thomson1980johnstone} to measure tidal currents in Johnstone Strait is smaller than the cross-sectional area of our northern boundary. Finally, the remaining constituents were added, starting with O$_1$, then S$_2$, and then in pairs P$_1$/N$_2$ and Q$_1$/K$_2$. Final adjustments were made by slightly varying amplitude and phase, evaluating the changes to M$_2$, K$_1$ and the ratios S$_2$/M$_2$, N$_2$/M$_2$, K$_2$/M$_2$, O$_1$/K$_1$, P$_1$/K$_1$, Q$_1$/K$_1$ and then choosing new values to move closer to observations by linear interpolation.
